---
title: "Week 09 - regularized regression/classification"
format:
  html: 
    embed-resources: true
---

# Learning goals

* Normal linear regression is meaningsless with k > n (more predictors than samples)
* Bias/Variance trade off. Overfitting is a problem with high dimensional data
* Use cross validation to estimate test error
* Try both lasso and ridge regression - and know the difference

# Data - phenotype prediction from blood metabolomics

We will work on public data where we try to predict two phenotypes of people - age and sex from different molecules in blood plasma.

The data is a mix of LCMS, GCMS and NMR data and it has all been combined here. The different names of the compounds are not super interesting for the purpose of this exercise.

## Data source

Metabolite patterns predicting sex and age in
participants of the Karlsruhe Metabolomics
and Nutrition (KarMeN) study

PLOS One

https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0183228

# Book labs

Before proceeding - go to the book labs and go through 

6.5.2 Ridge Regression and the Lasso

then come back here.

(Maybe you could just write your code here - of you like)

And if you feel for it, you are welcome to go through the first part of the labs as well (subset selection etc).

```{r}

# Beginning of chapter 6 labs

library(ISLR2)

head(Hitters)

Hitters <- na.omit(Hitters)

```

# Karmen analysis

## Loading data

```{r}

library(tidyverse)

df <- read_rds(file = "data/karmen_data.rds")

```

## Making a training/test split for all of us

To make it easier for all of us, we share some code so we save 20% of the data for validation (the same data for all of us)

```{r}

set.seed(0)

x <- model.matrix(age~., df)[, -1]
y <- df$age

df <- df %>%
  mutate(set = sample(x=c("Training set", "Test set"), size = n(), replace = TRUE, prob = c(0.80,0.20))) %>%
  select(set, everything())

train <- which(df$set=="Training set")
test <-  which(df$set=="Test set")

```

Now you have two vectors of row numbers called "train" and "test".

# Data inspection

## >Q1.1: How many predictors do we have in the data? (dimensions)
```{r}
dim(df)

#There are 442 predictors
```


--- 

## >Q1.2: How are the predictor values distributed? (scaling)

We will ignore any scaling requirements for now. But it's important that you can identify issues with data, e.g. are the data on different scales?

The quick and dirty check is to check the distribution of all the values combined and also log transformed.
```{r}
hist(as.matrix(df %>% select(-set)), 
     main="Distribution of Original Values", 
     xlab="Value",
     col="skyblue", 
     breaks=50)

hist(log10(as.matrix(df %>% select(-set)) + 1), 
     main="Log10-Transformed Distribution (+1 offset)", 
     xlab="Log10(Value + 1)",
     col="salmon", 
     breaks=50)
```


--- 

## >Q1.3: How is our response (age) distributed? (should we transform it?)

We should check if the dataset is balanced or looks weird.

Again, we will do nothing but we should check it.


--- 

## >Q1.4: Make a PCA plot of the data, with age as color.

Do you see a pattern/tendency in the data? 


--- 

# Ridge regression

## >Q2.1: Perform cross validated ridge regression on the training data, calculate the training and test RMSE.

There are three error metrics that are commonly used for evaluating and reporting the performance of a regression model; they are: Mean Squared Error (MSE). Root Mean Squared Error (RMSE). Mean Absolute Error (MAE).

Discuss what you expect BEFORE you do this. What do you expect of training and test RMSE?

Also, make plots with the predicted vs. the observed age for both sets.


--- 

## >Q2.2: Find to top 10 molecules with the largest effect sizes


Hint: again, check the book labs to get coefficients from the model.


--- 

# Lasso regression

## >Q3.1: Perform cross validated lasso regression on the training data, calculate the training and test RMSE.

Once you have the ridge regression up and running this should be pretty quick. Basically setting alpha=1 instead of alpha=0...


--- 

## >Q3.2: Find to top 10 molecules with the largest effect sizes

Also: count how many coefficients are not zero.

Hint: again, check the book labs to get coefficients from the model.


--- 

# Comparison of lasso and ridge

## >Q4.1: Which method had lowest test RMSE?

Discuss the training error (RMSE) differences between the two. What happened?

## >Q4.2: What does it tell you? 

So, think about if you expect all, some or few of the molecules to change with age. Is your expectation in line with what you observe?

## >Q4.3: What about non linearity and interactions ?

Do the chosen method allow for non-linear relations? What about interactions in the data?

(We will come back to this dataset later on).

## >Q4.3: What should we do? 

We do actually want to use data very similar to this find biomarkers we can use for estimating the age of blood spots from crime scenes. 

Which method would you prefer? Our goal is to find <25 target molecules, since we have super sensitive methods for measuring a few molecules with insane precision (targeted LCMS).

You can read a little about the project here (in Danish - you can probably get google to translate it): 

https://innovationsfonden.dk/da/i/historier/ny-retskemisk-metode-skal-afsloere

# Bonus

## >Q5: how robust are the error estimates for ridge and lasso if you have different train/test sizes?

So... basically I'm curious to see how much your estimates changes from run to run as well as with the size of the train/test set.

I the beginning of this exercise we set the training part of the data to be 80% and the test part to be 20%.

What do you think will happen if we use e.g. 50:50 ? (Then we have less training data but more test data.)

When we have MORE training data do you think the models becomes BETTER/WORSE at generalizing (test error)?

When we have LESS test data do you think the test error estimate is MORE/LESS variable from run to run?

If we had infinite traning and test data - what would you expect? (for this dataset)


--- 

