---
title: "Week 04 in Statistical Machine Learning in Bioinformatics"
subtitle: "classification I: logistic regression"
author: "Thomas Bataillon"
date: "Time Stamp last update: `r Sys.Date()`"
output:
  html_document:
    theme: readable
    code_folding: show
    toc: true
    toc_depth: 2
editor_options: 
  chunk_output_type: inline
---

# Goal of the session

-   Get familiarized with how over fitting in n\>p or n\<\<p works

-   Exploring the tcga dataset

-   Exploring logistic regression and its use as for predicting tissue type (Normal or Cancerous) in the tcga dataset

-   Gearing up for submitting your first set of predictions on cancer

# A fake dataset with no signal

```{r}
library(tidyverse)
library(ggthemes)
library(knitr)
theme_set(theme_minimal(base_size = 15, base_line_size = 1.5))

set.seed(42)
n_ind <- 300  # How many individuals in each class
n_features <- 10^4 # How many spurious features ( aka predictors)

fake_features <- rnorm(n = 2*n_ind*n_features) # 2*n_ind in total * n_features
fake_df <- as_tibble(matrix(data = fake_features, ncol = n_features))
dim(fake_df)
rm(fake_features) # not needed --> cleanup

reponse_binary <- as.factor(c(rep("Control", n_ind), rep("Case", n_ind)))
reponse_score <- rnorm(n = 2*n_ind, mean = 10, sd = 2)
length(reponse_score)
pseudo_df  <- tibble(reponse_binary, fake_df)

pseudo_df %>% select(1:6) %>% head() %>% kable("pandoc", digits = 2)

```

# A bit of tidyverse magic with `pivot_longer`

First reorganize, and lets pretend these features are gene expression data, typically expressed on log scale because gene expression can vary by 2-3 orders of magnitude.

```{r}
df_longer <- pivot_longer(data = pseudo_df,
                          cols =2:(n_features+1),names_to = "Feature")
dim(df_longer)
names(df_longer) <-c("response", "Feature", "log10_gene_expr")
head(df_longer) %>% knitr::kable(digits = 2)
```

Then summarize by gene:

```{r}
stats_byFeature_by_treatment <-df_longer %>%
  group_by(Feature,response) %>%
  summarize(n_obs = n(),   # just a check sample size
            mean = mean(log10_gene_expr),
            var = var(log10_gene_expr))

head(stats_byFeature_by_treatment) %>% knitr::kable(digits = 2)
```

# More tidyverse with `pivot_wider`

and this time pivot wider ... to build a series of summaries per features ...

```{r}
summary_features <-pivot_wider(data = stats_byFeature_by_treatment, 
            names_from = response, 
            values_from = c("n_obs","mean", "var"))

head(summary_features) %>% knitr::kable(digits = 2)

```

# A standardized difference approach - aka the `t-test`

This is the most celebrated statistic to look at differences between 2 groups

```{r}
summary_features  <- summary_features %>%
  mutate(mean_diff = mean_Case - mean_Control) %>%
  mutate(se_diff = sqrt( (var_Case/n_obs_Case) +(var_Control/n_obs_Control)) ) %>%
  mutate(t_welch = mean_diff / se_diff) %>%
  arrange(-abs(t_welch)) %>%
  {.}

knitr::kable(summary_features[1:10, c(1,8,10)], col.names = c("gene", "Obs_diff", "t_obs"))

```

# The distribution of `t_obs` (under zero effect aka $H_0$)

```{r}
ggplot(data = summary_features, aes(x=t_welch))+ 
  geom_histogram(bins = 20, fill="cornflowerblue")+  # <--my kind of blue :)
  xlab("standardized difference in Means vs. Controls")+
  theme_minimal(base_size = 15)+
  NULL


```

# Cherry picking: the best 2

```{r}
pseudo_df <- pseudo_df %>%
  mutate(logistic01 = ifelse(reponse_binary == "Case", 0, 1))
 
table(pseudo_df$logistic01)

ggplot(pseudo_df, aes(x=V1608, y = V3630, color = reponse_binary)) +
  geom_point(size=2, alpha=0.85)+
  theme_minimal(base_size = 15)+
  theme(legend.position = "bottom")+
  NULL

```

# Predicting after cherry picking 2/n

```{r}


m1 <- glm(data = pseudo_df, formula = logistic01 ~ V1608 + V3630 ,  family = "binomial" )
m0 <- glm(data = pseudo_df, formula = logistic01 ~ 1,  family = "binomial" )
(1 - m1$deviance/m0$deviance)

pseudo_df$magic <-m1$fitted.values
ggplot(data = pseudo_df, aes(x=magic, fill = reponse_binary)) + 
  geom_dotplot(binwidth = 0.0150)+ 
  scale_fill_colorblind()+
  xlab("2 logistic regression predictors") +
  NULL


```

# Predicting after cherry picking 3/n

## \>Q00 explore how much prediction can be achieved using 5-10 carefully chosen predictors

```{r}
m1 <- glm(data = pseudo_df, formula = logistic01 ~ V1608 + V3630 + V2976 + V5864 + V1709 + V1777 + V148 + V9460 + V287 + V7702 ,  family = "binomial" )
m0 <- glm(data = pseudo_df, formula = logistic01 ~ 1,  family = "binomial" )
(1 - m1$deviance/m0$deviance)
#pseudo_df$reponse_binary <- factor(pseudo_df$reponse_binary, levels = c("Control", "Case"))
pseudo_df$magic <-m1$fitted.values
ggplot(data = pseudo_df, aes(x=magic, fill = reponse_binary)) + 
  geom_dotplot(binwidth = 0.0150)+ 
  xlab("10 logistic regression predictors") +
  NULL
```

```{r}
# Splitting the data: 80% for training, 20% for testing
set.seed(123)
train_idx <- sample(1:nrow(pseudo_df), size = 0.8 * nrow(pseudo_df))
train_data <- pseudo_df[train_idx, ]
test_data  <- pseudo_df[-train_idx, ]

# Training the model ONLY using the training data
m_train <- glm(data = train_data, 
               formula = logistic01 ~ V1608 + V3630 + V2976 + V5864 + V1709 + V1777 + V148 + V9460 + V287 + V7702, 
               family = "binomial")

# Prediction outcomes for the TEST data
test_predictions <- predict(m_train, newdata = test_data, type = "response")

# Plot the TEST results
test_results <- tibble(prob = test_predictions, real = test_data$reponse_binary)

ggplot(test_results, aes(x = prob, fill = real)) +
  geom_dotplot(binwidth = 0.05) +
  scale_fill_colorblind() +
  labs(title = "Predictions on UNSEEN data (Test Set)", 
       x = "Predicted Probability", 
       fill = "Actual Status") +
  theme_minimal()
```

# Background on the TCGA data

-   The dataset we work on for the next weeks is a slice of the original tcga gene expression dataset. More precisely, here we use a random subset of 4000 genes out of ca 18000 genes from the initial dataset.

-   Each individual sample (`rowid`) has a matching metadata where a sample type (`reponse`: Normal / Cancer) are given along with a `tissue` type ( for the training part of the data)

-   The gene expression data comprising 4000 genes where log10(gene expression) is provided + 3 first principal components (aka `pc1` `pc2` `pc3`)

-   About the PCs: `pc1` and `pc2` these 2 synthetic variables are orthogonal and account for ca 20% of the total variation in gene expression in the 4000 genes we use for prediction

IMPORTANT: for some samples (1172 of them) the nature of the tissue (`reponse`) is not known (`NA`) and should be predicted (we will try various methods to do so this week and in the coming weeks)

# 1. Loading the data:

We will load the dataset The file is a tible saved in R's rds format (with gz compression). Rds is the fastest format for writing and reading.

Each row is one biological sample (one persons tissue)

It contains info on the sample and pc1, pc2 and log10(1 + gene expression) data.

You will se that the response is unknown/missing for some of the samples (\~20 percent of the data).

The response column is recoded for the logictic regression:\
\* 0 = `Normal`, \* 1 = `Tumor`

```{r message=FALSE, warning=FALSE}
library(tidyverse)
getwd()
tcgadf <- read_rds("data/TCGA_cancer_classification/miniTCGA.3349x4006.rds") 

dim(tcgadf) # check this should be 3349 4007
names(tcgadf)[1:30]

tcgadf %>%
  group_by(response) %>%
  summarise(observations=n()) %>%
  knitr::kable()


tcgadf <- tcgadf %>%
  mutate(sampletype=response, 
         response=ifelse(test= response=="Tumor", yes = 1, no = 0 ))

tcgadf %>%
  group_by(response) %>%
  summarise(observations=n()) %>%
  knitr::kable()


```

## \> Q0: First, think about the "structure" of the data:

-   What are the predictors ?\
    The predictors (also known as independent variables or features) are the **gene expression levels** and the principal components (**PC1, PC2**).

    -   In this dataset, each column (except for `response` and `sampletype`) represents a genomic feature, specifically the $log_{10}(1 + \text{gene expression})$ values for thousands of different genes.

-   what is the response variable you are interested in ?

    The response variable is **`response`**.

    -   It is a **binary categorical variable** that has been recoded into numerical values for the logistic regression: **0** for `Normal` tissue and **1** for `Tumor` tissue.

-   Are we in the n\>\>p, n\>p, n\~p, n\<p or n\<\<p "domain" ?

    We are in the $n \ll p$ domain (where $n$ is much smaller than $p$).

# 2. vizualizing the TCGA dataset

## \> Q1: Do some simple graphical visualizations,

Explore to see how 2 gene expression levels or 2 pc of gene expression differ by Tissue and SampleType

```{r}
tcgadf_no_na <- subset(tcgadf, !is.na(tissue))

tcgadf_clean <- tcgadf_no_na %>%
  mutate(sample_type = case_when(
    response == 0 ~ "Normal",
    response == 1 ~ "Tumor"
  ))

ggplot(tcgadf_clean, aes(x = tissue, y = pc1, fill = sample_type)) +
  geom_boxplot()+
  scale_fill_manual(values = c("Normal" = "green", "Tumor" = "red")) +
  labs(title = "PC1 by tissue")

ggplot(tcgadf_clean, aes(x = tissue, y = pc2, fill = sample_type)) +
  geom_boxplot()+
  scale_fill_manual(values = c("Normal" = "green", "Tumor" = "red")) +
  labs(title = "PC1 by tissue")
```

------------------------------------------------------------------------

> Q: draw "by eye" the boundaries between types (Normal / cancer) and tissue type

# 3. Formatting preparing the data for using the classification methods

## \> Q2: build a training set by filtering the data so you only have samples with SampleType known

```{r}
# 1. Create the training set by removing rows with unknown SampleType
training_set <- tcgadf_clean %>%
  filter(!is.na(sample_type))

# 2. Check how many samples we have left
nrow(training_set)

# 3. See the distribution of Normal vs Tumor
table(training_set$sample_type)
```

------------------------------------------------------------------------

# Part 4: Trying out classification methods

Try out the logistic regression on the tcga dataset. If you are unsure about the R function glm() that is implementing logistic regression, then lookup the R labs of the classification chapter.

## \> Q3: Can you identify specific genes that are good at discriminating between types (normal / Cancer) overall and within a specific tissue ?

```{r}
library(tidyr)
library(dplyr)

long_data <- training_set %>%
  pivot_longer(
    cols = -any_of(c("rowid", "response", "tissue", "pc1", "pc2", "pc3", "sample_type", "sampletype")), 
    names_to = "gene",
    values_to = "expression"
  )

summary_overall <- long_data %>%
  group_by(gene) %>%
  summarise(
    n_obs_Case = sum(response == 1, na.rm = TRUE),
    n_obs_Control = sum(response == 0, na.rm = TRUE),
    mean_Case = mean(expression[response == 1], na.rm = TRUE),
    mean_Control = mean(expression[response == 0], na.rm = TRUE),
    var_Case = var(expression[response == 1], na.rm = TRUE),
    var_Control = var(expression[response == 0], na.rm = TRUE)
  ) %>%
  mutate(mean_diff = mean_Case - mean_Control) %>%
  mutate(se_diff = sqrt( (var_Case/n_obs_Case) + (var_Control/n_obs_Control) )) %>%
  mutate(t_welch = mean_diff / se_diff) %>%
  arrange(desc(abs(t_welch))) # Ordena por el valor absoluto de t_welch

knitr::kable(summary_overall[1:10, c("gene", "mean_diff", "t_welch")], 
             col.names = c("Gene", "Obs_diff", "t_obs"))
```

## \> Q4: Explore how pc1 and pc2 can effectively summarize the info of 500 gene expression measurements

```{r}
ggplot(tcgadf_clean, aes(x = pc1, y = pc2, color = as.factor(response))) +
  geom_point(alpha = 0.6) +
  labs(title = "Dimensionality Reduction: PC1 vs PC2",
       subtitle = "Summarizing gene expression to separate Normal (0) vs Tumor (1)",
       x = "Principal Component 1", 
       y = "Principal Component 2", 
       color = "Tissue Type") +
  theme_minimal() +
  scale_color_manual(values = c("0" = "blue", "1" = "red"))

# Fit a logistic regression using only the first two principal components
pc_model <- glm(response ~ pc1 + pc2, data = tcgadf_clean, family = "binomial")

# View the summary to see if pc1 and pc2 are statistically significant predictors
summary(pc_model)
```

This statistical output confirms that your model is highly successful: the extremely low p-values `(<2e-``16`) prove that PC1 and PC2 are both highly significant predictors that effectively capture the biological signal from the original 500 genes. The substantial drop in deviance (from 1481.5 to 1036.9) indicates that these two synthetic variables drastically improve the model's ability to distinguish between normal and cancerous tissue, providing a robust classification without the risk of overfitting thousands of individual predictors.

## \> Q5: try and combine up to max 3 predictors to make your classification ..

To start, by limiting the number of predictors, the model is more easy to "interpret". Later on we will go "bananas" and increase the number of predictors if it is sensible to do so (see cross validation and model selection inthe coming weeks )

------------------------------------------------------------------------

Hint: you could try to use the `broom` package and use that to organize the results of different models.

Note that the statistics computed by glance are different for `glm()` objects than for `lm()` (e.g. deviance rather than R\^2 are reported). make sure you understand these terms and if in doubt ask your TA :)

```{r}
library(broom)

# glm with the top 3 genes from Q3
gene_model_3 <- glm(response ~ ANGPTL1.9068 + NR3C2.4306 + TMEM220.388335, 
                    data = tcgadf_clean, 
                    family = "binomial")

# Use tidy() to look at the coefficients and p-values for each gene
tidy(gene_model_3)

# Use glance() to look at the overall model fit (Deviance and AIC)
glance(gene_model_3)
```

------------------------------------------------------------------------

The first table confirms that three chosen genes are incredibly strong, independent predictors of tissue type. The extremely small **p-values** (all well below 0.05) indicate that each gene contributes significant predictive power to the model. Additionally, the negative **estimates** (coefficients) for all three genes mean that higher expression levels of ANGPTL1, NR3C2, and TMEM220 are strongly associated with a lower probability of the tissue being cancerous. Essentially, these genes are highly expressed in normal tissue and appear to be down-regulated or "turned off" in tumors.

The second table reveals that this targeted 3-gene model drastically outperforms your previous Principal Component model. The **residual deviance** dropped all the way down to 539.9 (compared to 1036.9 in the PC model), and the **AIC** is nearly cut in half to 547.9 (down from 1042.9). Because lower values in both of these metrics indicate a better model fit with less error, this proves that selecting just a few highly informative, specific genes can provide a much sharper and more accurate classification than summarizing hundreds of genes into broader synthetic variables like PC1 and PC2.

# Part 5. looking at model fit

##\> Q6: Make some visual presentation of the glm fit

use `ggplot()` to produce the equivalent of figure 4.2 of the book for the logistic regression model you chose.

```{r}
# First, we reshape the data using pivot_longer so ggplot can read it easily
plot_data_3genes <- tcgadf_clean %>%
  select(response, ANGPTL1.9068, NR3C2.4306, TMEM220.388335) %>%
  pivot_longer(cols = -response, names_to = "Gene", values_to = "Expression")

# Now we plot it, adding 'facet_wrap' to split it into 3 panels
ggplot(plot_data_3genes, aes(x = Expression, y = response)) +
  geom_point(shape = "|", color = "orange", size = 4, alpha = 0.5) +
  geom_smooth(method = "glm", method.args = list(family = "binomial"), 
              color = "dodgerblue", se = FALSE) +
  
  facet_wrap(~ Gene, scales = "free_x") + 
  
  labs(title = "Logistic Regression Fits for Top 3 Genes",
       x = "Gene Expression Level",
       y = "Probability of Tumor") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray") +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray") +
  theme_classic()
```

Hint: if you still struggle with all the arcane options of `ggplot()` try this post :)

<https://stackoverflow.com/questions/47080625/plotting-predictions-from-a-logistic-regression>

------------------------------------------------------------------------

# Well done! TOu are almost ready to make your first round of predictions ?? !!
